DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-11-02 2015CompleteDerivation/completeDerivation.tex
Another pass at quantum prelude

While in principle, we could simply introduce assumption \ref{ass:irreducibility} and work our way to quantum mechanics, there is a non insignificant chance of being misconstrued. Different people come with different background, to the point that the same words trigger different intuitions and expectations. Therefore we will first ground in classical mechanics the intuition that will be needed later, making the step into quantum mechanics a gentler, and hopefully less confusing, one. It will also help us in showing that classical and quantum mechanics are not, after all, so different.

We will see that, as a first approximation, quantum particles are more keenly described by classical composite states (i.e. a distribution in phase space) as they are made of a finite amount of material. In fact, classical distributions already possess qualitative properties that are typically associated with quantum systems such as a classical analogue of the uncertainty principle.

@@

As we defined them in previous sections, classical particles should not be understood as point-like objects of finite mass. It was only by assuming they were infinitesimal parts of an invariant density distribution that gave us conjugate pairs of state variables. Hamiltonian mechanics would not have been justified otherwise.

Any finite amount of classical material is described by a distribution. Therefore, to a first approximation, an electron, which is a finite amount of charge/mass, must be described by a distribution as well. This may seem preposterous at first, but we contend that it is no more preposterous than to think it point-like, described by a probability distribution that self-interacts. In fact, the idea of an electron being a distribution of material is already conceptually more compatible with two issues.

The first one is that an electron does not seem to have a well defined position or momentum. If we misidentify the electron with a classical particle, this would indeed be worth of note and puzzlement. But if the electron is a classical distribution, there is not issue. The earth is both under Milano and New York. Your right index finger is not, hopefully, at the same position of your left eye. In fact, based on what we saw before, any object of finite mass is a distribution over a region, and discontinuous distributions are ruled out by \ref{continuous_map}. An electron is an object of finite mass: why should it be any different? In fact, if one claimed an electron to be point-like, he would be claiming an electron behaves in a different way than all other physical objects of finite mass (which are the only ones we experimentally work with). Shouldn't \emph{he} have the burden of proof?

The other issue is particle-wave duality. If the spread of the distribution is narrow enough for the problem at hand, the motion can be described by the position and center of mass, making it behave like a classical particle. If the spread is not negligible (e.g. it encounters obstacles of a smaller scale) then we have to use the full distribution. Therefore, even in this classical approximation, the electron would have this dual nature.

@@

In fact, we can construct an infinitesimal object that is not elementary. Suppose we have a sphere with dots moving on its surface. All their dynamics can be expressed by their polar and azimuthal angles. If we scale the sphere, the dynamics stays the same. We can shrink the whole sphere to a point: it is now a point-like object that is not elementary, as there is motion on its surface detectable along different angle of incidence.

@@

Suppose we have our classical electron described by a finite amount of classical material distributed over a single d.o.f. $\int \rho_\mathcal{c} dq \wedge dp = 1$ where $\rho_\mathcal{c}$ is the normalized distribution. In this context, we will call \emph{fragment} an infinitesimal part of our classical electron. The Shannon entropy associated with the distribution, which can be though as the amount of information required to identify a fragment given the distribution, is finite as well.\footnote{To be precise, Shannon's differential entropy is actually the number of bits needed relative to a uniform distribution of unitary range on all axis. The intuitive meaning, though, is the same.} $I(\rho_\mathcal{c}) = \int \rho_\mathcal{c} \log (\rho_\mathcal{c})  dq \wedge dp = I_0$. TODO: prove in appendix.

As the state evolves into its final state, the amount of material remains the same: $\int \rho_{\hat{\mathcal{c}}} dq \wedge dp = \int \rho_{\mathcal{c}} dq \wedge dp$. The final entropy will also remain the same: $I(\rho_{\hat{\mathcal{c}}}) = I_0$. Mathematically because $\rho_\mathcal{c} \log (\rho_\mathcal{c})$, like the distribution itself, is just another integrable function. Physically, because the amount of information required to identify a fragment given the distribution has to remain the same under deterministic and reversible evolution. Therefore, in the same way that each fragment is constrained to move along states at equal energy, the composite state is constrained to move along distributions at equal entropy.

We now ask the following question: once we fix the entropy, what is the relationship between the spread of the marginal distributions along $q$ and $p$? Using Lagrange multipliers, we fix the total amount of material to $1$ and the total entropy $I_0$ and find the distribution that minimizes the the product of the variance $\sigma_q^2 \sigma_p^2 = \int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp$.
\begin{align*}
L = &\int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp \\
&+ \lambda_1(\int \rho dqdp - 1)\\
&+ \lambda_2(- \int \rho \ln \rho \, dqdp - I_0)\\
\delta L = &\int \delta \rho [(q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\ &+ \lambda_1 + \lambda_2 \ln \rho + \lambda_2 ] dqdp = 0 \\
\rho = &\frac{1}{ \sqrt{4 \pi ^2\sigma_q^2\sigma_p^2}} e^{-\frac{(q-\mu_q)^2(p-\mu_p)^2}{4\sigma_q^2\sigma_p^2}} \\
= &\frac{1}{e^{I_0 - 1}} e^{-\frac{\pi ^2(q-\mu_q)^2(p-\mu_p)^2}{\exp (2(I_0 - 1))}}
\end{align*}
We find that the distribution that minimizes the spread is the product of two independent Gaussians. Recall that in quantum mechanics, the Gaussian wave packet is the state that minimizes uncertainty: this is the classical analogue.

@@

Suppose, though, the evolution of the distribution is deterministc and reversible while each fragment is undergoing non deterministic motion. Suppose $\rho$ is a distribution at equilibrium of such motion. While each fragment moves randomly, the overall distribution $\rho$ stays the same: the non deterministic motion is effectively just reshuffling the fragments within the same distribution. The value of the entropy of the distribution, and therefore its spread, is set by the non deterministic motion of the particles: the stronger it is, the greater the entropy.

Suppose now the classical electron undergoes quasi-static evolution. At each instant we can assume $\rho$ to be at equilibrium. At each instant the fragments are reshuffled but the amount of entropy of the distribution remains the same. The overall evolution is still Hamiltonian, but the trajectories predicted by the framework are now just the average motion of the fragments.

If we assumed that all our classical electrons were interacting with the same background that caused all fragments to exhibit a random walk of the same magnitude, as defined by the information entropy, then the spread of all distributions for all material would satisfy the same inequality. This gives us the classical uncertainty principle: $\sigma_q\sigma_p \geq \exp (I_\varnothing - 1) / 2 \pi $ where $I_\varnothing$ is the entropy associated with the background.

Assuming such a background is not far fetched. In fact, it serves as a remainder of the issues mentioned in assumption \ref{ass:infinitesimal_reducibility}: as determinism and reversibility at all scales conceptually does not work, there exists a minimum constant interaction: the system cannot be perfectly isolated. If the entropy of our distribution is comparable to the entropy introduced by the background, we are likely in the range where classical mechanics is not valid as the random motion of each fragment is not negligible.

The generalization to multiple d.o.f. is outside of the scope of this work. Note, though, that this result is similar in spirit to the Gromov non-squeezing theorem and the symplectic camel, which are already generalized to multiple d.o.f.. It is the link with information theory that gives us further physical insight and is of particular interest given the strong connection between thermodynamics and Shannon's entropy. This is something to be developed in a future work.

>>>>

Before introducing assumption \ref{ass:irreducibility} and deriving quantum mechanics as its consequence, we first give a semi-classical account of the conceptual model we will be using. This allows to build the intuition more gradually and to show that classical and quantum mechanics are not as different as one may first imagine.

We will see that, as a first approximation, quantum particles are more keenly described by classical composite states (i.e. a distribution in phase space) like any other object made of a finite amount of material. We'll also see that classical distributions already possess qualitative properties that are typically associated with quantum systems, including a classical analogue of the uncertainty principle.

@@

As we defined them in previous sections, classical particles should not be understood as point-like objects of finite mass. It was their being infinitesimal and spread around a small neighborhood that gave us conjugate pairs of state variables and Hamiltonian evolution. Any finite amount of classical material, instead, is described by a distribution.

If we take a quantum particle, say an electron, it has a finite amount of charge/mass/energy. Therefore, to a first approximation, it must be described by a distribution as well. This may, to some, seem preposterous at first: an electron is not a point particle. But we contend that it is no more preposterous than to think it point-like, described by a probability distribution that self-interacts. In fact, the idea of an electron being a distribution of material is already conceptually more compatible with two issues.

The first one is that an electron does not seem to have a well defined position or momentum. If we misidentify the electron with a classical particle, this would indeed be worth of note and puzzlement. But if the electron is a classical distribution, there is no issue. The earth is both under Milano and New York. Your right index finger is not, hopefully, at the same position of your left eye. In fact, based on what we saw before, any object of finite mass is a distribution over a region, and discontinuous distributions are ruled out by \ref{continuous_map}. An electron is an object of finite mass: why should it be any different? In fact, if one claimed an electron to be point-like, he would be claiming an electron behaves in a different way than all other physical objects of finite mass/energy (which are the only ones we experimentally work with). Shouldn't \emph{he} have the burden of proof?

The other issue is particle-wave duality. If the spread of the distribution is narrow enough for the problem at hand, the motion can be described by just the position and momentum of the center of mass, making it behave like a classical particle. If the spread is not negligible (e.g. it encounters obstacles of a smaller scale) then we have to use the full distribution. Therefore, even in this classical approximation, the electron would have this dual nature depending on the problem.

@@

In fact, we can conceptually construct an infinitesimal object that is not elementary. Suppose we have dots moving on the surface of a sphere. Their motion can be expressed in terms of their polar and azimuthal angles. If we shrink the sphere, the dynamics and its description stays the same. We can collapse the whole sphere to a point: it is now a point-like object that is not elementary, as there is motion detectable along different angle of incidence.

@@

Suppose we have our classical electron described by a finite amount of classical material distributed over a single d.o.f. $\int \rho_\mathcal{c} dq \wedge dp = 1$ where $\rho_\mathcal{c}$ is the normalized distribution. In this context, we will call \emph{fragment} an infinitesimal part of our classical electron. The Shannon entropy associated with the distribution is finite as well. $I(\rho_\mathcal{c}) = - \int \rho_\mathcal{c} \log (\rho_\mathcal{c})  dq \wedge dp = I_0$. Intuitively, this value can be though as the amount of information required to identify a fragment within the distribution.\footnote{To be precise, Shannon's differential entropy is actually the number of bits required by the given distribution relative to a uniform distribution of unitary range on all axis. The intuitive meaning, though, is the same.} TODO: prove in appendix.

As the state evolves into its final state, the amount of material remains the same: $\int \rho_{\hat{\mathcal{c}}} dq \wedge dp = \int \rho_{\mathcal{c}} dq \wedge dp$. The final entropy will also remain the same: $I(\rho_{\hat{\mathcal{c}}}) = I_0$. Mathematically because $-\rho_\mathcal{c} \log (\rho_\mathcal{c})$, like the distribution itself, is just another integrable function whose integral is conserved by Hamiltonian evolution. Physically, because the amount of information required to identify a fragment given the distribution has to remain the same under deterministic and reversible evolution. Therefore, in the same way that each fragment is constrained to move through states at equal energy, the composite state is constrained to move through distributions at equal entropy.

We now ask the following question: once we fix the entropy, what is the relationship between the spread of the marginal distributions along $q$ and $p$? Using Lagrange multipliers, we fix the total amount of material to $1$ and the total entropy $I_0$ and find the distribution that minimizes the the product of the variance $\sigma_q^2 \sigma_p^2 = \int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp$.
\begin{align*}
L = &\int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp \\
&+ \lambda_1(\int \rho dqdp - 1)\\
&+ \lambda_2(- \int \rho \ln \rho \, dqdp - I_0)\\
\delta L = &\int \delta \rho [(q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\ &+ \lambda_1 - \lambda_2 \ln \rho - \lambda_2 ] dqdp = 0 \\
\lambda_2 \ln \rho = &\lambda_1 - \lambda_2 + (q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\
\rho = &e^{\frac{\lambda_1 - \lambda_2}{\lambda_2}}e^{\frac{(q-\mu_q)^2 \sigma_p^2}{\lambda_2}}e^{\frac{\sigma_q^2 (p-\mu_p)^2}{\lambda_2}}\\
\rho = &\frac{1}{ 2 \pi \sigma_q \sigma_p} e^{-\frac{(q-\mu_q)^2}{2\sigma_q^2}} e^{-\frac{(p-\mu_p)^2}{2\sigma_p^2}} \\
I_0 = &\ln (2\sigma_q\sigma_p\pi) + 1
\end{align*}
(TODO: move the details of the math) We find that the distribution that minimizes the spread is the product of two independent Gaussians. Recall that in quantum mechanics, the Gaussian wave packet is the state that minimizes uncertainty: this is the classical analogue.

@@

Suppose, though, that while the evolution of the distribution of our electron electron as a whole is deterministc and reversible, each fragment is undergoing non deterministic motion. Suppose $\rho$ is a distribution at equilibrium of such motion. While each fragment moves randomly, the overall distribution $\rho$ stays the same: the non deterministic motion is effectively just reshuffling the fragments within the same distribution. The value of the entropy of the distribution, and therefore its spread, is set by the non deterministic motion of the particles: the stronger it is, the greater the entropy.

Suppose now our classical electron undergoes quasi-static evolution. At each instant we can assume $\rho$ to be at equilibrium. At each instant the fragments are reshuffled but the amount of entropy of the distribution remains the same. The overall evolution is still Hamiltonian, but the trajectories predicted by the framework are now just the average motion of the fragments.

If we assumed that all our classical electrons were interacting with the same background, all fragments would exhibit a random walk of the same magnitude as defined by the information entropy. The spread of all distributions for all material would satisfy the same inequality. This gives us the classical uncertainty principle: $\sigma_q\sigma_p \geq \exp (I_\varnothing - 1) / 2 \pi $ where $I_\varnothing$ is the entropy associated with the background.

Assuming such a background is not far fetched. We could justify it to address the issues mentioned in assumption \ref{ass:infinitesimal_reducibility}: as determinism and reversibility at all scales conceptually does not work, we posit a minimum constant interaction: the system cannot be perfectly isolated. If the entropy of our distribution is comparable to the entropy introduced by the background, we are likely in the range where classical mechanics is not valid as the random motion of each fragment is not negligible.

The generalization to multiple d.o.f. is outside of the scope of this work. Note, though, that this result is similar in spirit to the Gromov non-squeezing theorem and the symplectic camel, which are already generalized to multiple d.o.f.. It is the link with information theory that gives us further physical insight and is of particular interest given the strong connection between thermodynamics and Shannon's entropy.
